---
title: "Literate Analysis"
author: "AT"
format: html
---

## Introduction

### About these data
This data set contains North Pole Environmental Observatory Bottle Chemistry data. This data set includes diverse measurements of seawater chemistry along with supporting conductivity, temperature, and depth (CTD) data from the Arctic Ocean near the North Pole (downloaded 6/26/23 from [this link](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X))

### Setup
```{r}
library(readr)
```

### Reading data
```{r}
#| warning: false
#| message: false

bg_chem <- read_csv("data/BGchem2008data.csv")
```


Exploring data

```{r}
#| eval: false
#| echo: false

colnames(bg_chem)

str(bg_chem)

head(bg_chem)

summary(bg_chem)

unique(bg_chem$Date)
```


## Analysis

### Calculate summary stats
```{r}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)
```

### Calculate mean Redfield ratio
```{r}
ratio <- (sum(nitrate, nitrite, amm)) / phos
```

### Plot Redfield ratio
```{r}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```


## Conclusion
The Redfield ratio for this dataset is approximately `r round(ratio)`

